Pleiotropy of gene regulatory variants increases with the number of regulated genes, active cell types and tissues and bound transcription factors

### 2022-11-17 ms/ms03_results.tex
MS

I found eQTL colocalizations for 3,849 or 36$\%$ of the leading GWAS variants with 2,348 eQTL genes and 127 eQTL samples.
% sqlite
% select count(distinct tophits.rsid) from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.`PP.H4.abf`>=0.8; 3849
% alternative: select count(distinct tophits.rsid) from tophits, coloc where coloc.rsid=tophits.rsid and coloc.`PP.H4.abf`>=0.8; 3849
% select count(distinct coloc.egene) from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.`PP.H4.abf`>=0.8
% select count(distinct coloc.eqtl_id) from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.`PP.H4.abf`>=0.8

@@

This pipeline has resulted in 103,551 variants from 246 GWAS that colocalized (PP.H4.abf$\geq$0.8) with at least one eQTL.
%select count(*) from (select DISTINCT gwas_id FROM coloc where `PP.H4.abf`>=0.8);
%select count(*) from (select DISTINCT rsid FROM coloc where `PP.H4.abf`>=0.8);
%
I have developed a web application where I expose 116,736 variants from 246 GWAS that colocalize with at least one eQTL at a lower threshold value PP.H4.abf$\geq$0.5.
% select count(*) from (select DISTINCT rsid FROM coloc where `PP.H4.abf`>=0.5);
% select count(*) from (select DISTINCT gwas_id FROM coloc where `PP.H4.abf`>=0.5);

@@

These category annotations were used to investigate the variant pleiotropy for GWAS traits.
%
Colocalized eQTL/GWAS variants were classified according to the number of GWAS categories they belong to (Table \ref{tab:pleitropic_variants} and Supplementary table 3).
%
These aggregated variants were highly pleiotropic.
%
For instance the variant rs2522051 is located upstream of the IRF1 gene in cytoband 5q31.1 and it is involved in six classes of traits, namely,
allergy, autoimmune diseases, cancer of breast, circulatory system diseases, hypertension and respiratory system diseases
(Table \ref{tab:pleitropic_variants} and Supplementary table 3).
%
This variant rs2522051 is active in 23 classes of cell type and tissues including adipose, brain, digestive cells, immune cells and sexual organs.
%
This variant rs2522051 regulates several genes such as IL13, PDLIM4 and RAD50.

\subsubsection*{What is the function of pleiotropic variants?}

%Then I evaluated the function of pleiotropic variants with 5, 4, 3 and 2 GWAS categories using gene ontology analysis.
%%
%I analyzed the ontology of eGenes with 5, 4, 3 and 2 GWAS categories using the DAVID web services \citep{2008.Lempicki.Huang,2008.Lempicki.Huang.NucleicAcidsResearch}.
%%
%I found that eGenes of pleiotropic variants with 3 and 2 GWAS categories were significantly enriched in functions of the immune system (Fig. \ref{s:results}{fig:geneontology}).
%For instance, eGenes in the cytoband 5q31.1 include IRF1 and IL4, which are important factors of the immune system.
%%
%IRF1 is a response protein to the presence of virus and oncogenic proteins and IL4 is required to stimulate proliferation of activated B and T-cells.
%%
%Variant rs17622656 is located in the intron of the IRF1 gene whereas variants rs736801 and rs2522051 are located upstream and downstream of the IRF1 gene, respectively.
%%
%These variants are associated with allergy, asthma, cardiovascular diseases, hypertension and ulcerative colitis and are active in a large number of tissues such as arteries, blood, brain, breast, digestive and immune system, muscle, skin, ovary, testis and thyroid.

%Variants of the 3q23 cytoband are involved in allergy, cancer and height traits (Supplementary table 3).
%%
%Interestingly all these variants target the eGene ZBTB3.
%%
%ZBTB3 is a transcription factor that controls pro-inflammatory factors such as IRF5 (10.1016/j.jaut.2016.08.001).
%%
%These variants are active in tissues such as blood, esophagus, immune cell, LCLs and muscle (Supplementary table 3).

%%11q13.5
%The variants in cytoband 11q13.5 are involved in allergy and autoimmune disease (Supplementary table 3).
%%
%All these variants target the eGene EMSY-DT, which is a long noncoding RNA upstream of EMSY.
%%
%These variants are active in LCL, Skin and iPSC (Supplementary table 3).
%
%In summary, I have identified pleiotropic variants with gene regulatory function.
%%
%Some of these loci are strongly involved in the immune system.

@@

Pleiotropic variants are aggregated in genomic regions.
%
Therefore I computed regions that include a large number of pleiotropic variants (See Methods) (Table \ref{tab:pleiotropic_regions}).
%
I found 13 regions with 5 or more GWAS classes, 30 regions with 4 or more GWAS classes and 80 regions with 3 or more GWAS.
%
75\% of regions are shorter than 100 kb, 15\% of regions are between 100 kb and 200 kb, and less than 5\% are larger than 200 kb (Fig. \ref{fig:pleiotropy_region_distribution}a).

The most pleiotropic region is 12:111,395,984-111,645,358 in cytoband 12q24.12 that we discussed above (Table \ref{tab:pleiotropic_regions}).
%
The largest region is 11:13,260,511-17,396,930 in cytoband 11p15.2 with a length of 4,136,419 bp and variants involved in cardiovascular and hypertension traits (Supplementary table 5).
%
The second largest region is 2:187,235,912-191,066,738 in cytoband 2q32.2 with a length of 3,830,826 bp and variants involved in cardiovascular, hypertension and autoimmune traits  (Supplementary table 5).

>>>>

My pipeline uncovered eQTL colocalizations for 3,849 or 36$\%$ of the leading GWAS variants with 2,348 eQTL genes and 127 eQTL samples.
% sqlite
% select count(distinct tophits.rsid) from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.`PP.H4.abf`>=0.8; 3849
% alternative: select count(distinct tophits.rsid) from tophits, coloc where coloc.rsid=tophits.rsid and coloc.`PP.H4.abf`>=0.8; 3849
% select count(distinct coloc.egene) from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.`PP.H4.abf`>=0.8
% select count(distinct coloc.eqtl_id) from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.`PP.H4.abf`>=0.8

@@

This pipeline resulted in 103,551 variants from 246 GWAS that colocalized (PP.H4.abf$\geq$0.8) with at least one eQTL.
%select count(*) from (select DISTINCT gwas_id FROM coloc where `PP.H4.abf`>=0.8);
%select count(*) from (select DISTINCT rsid FROM coloc where `PP.H4.abf`>=0.8);
%
I have also developed a web application where I expose 116,736 variants from 246 GWAS that colocalize with at least one eQTL at a lower threshold value PP.H4.abf$\geq$0.5.
% select count(*) from (select DISTINCT rsid FROM coloc where `PP.H4.abf`>=0.5);
% select count(*) from (select DISTINCT gwas_id FROM coloc where `PP.H4.abf`>=0.5);

@@

These class annotations were used to investigate the variant pleiotropy.
%
Colocalized eQTL/GWAS variants were classified according to the number of GWAS categories they belong to (Table \ref{tab:pleitropic_variants} and Supplementary table 3).
%
The aggregated variants are highly pleiotropic.
%
For instance the variant rs2522051 is located downstream of the IRF1 gene in cytoband 5q31.1 and it is involved in six classes of traits, namely,
allergy, autoimmune diseases, cancer of breast, circulatory system diseases, hypertension, and respiratory system diseases
(Table \ref{tab:pleitropic_variants} and Supplementary table 3).
%
This variant rs2522051 is active in 23 classes of cell types and tissues including adipose, brain, digestive cells, immune cells, and sexual organs.
%
This variant rs2522051 regulates several genes such as IL13, PDLIM4 and RAD50.

@@

Pleiotropic variants are aggregated in genomic regions, for instance, in cytobands 3q23, 5q31.1, 9p21.3, 15q24.1, and 19q13.33 for the variants involved on five or more classes (Supplementary table 3).
%
Therefore I computed regions that include a large number of pleiotropic variants (See Methods) (Table \ref{tab:pleiotropic_regions} and Supplementary table 4).
%
I found 13 regions with 5 or more trait classes, 30 regions with 4 or more trait classes and 80 regions with 3 or more trait classes (Supplementary table 4).
%
% TODO verify this plot, is plot 2b used somewhere?
75\% of regions involved in two or more trait classes are shorter than 100 kb, 15\% of regions are between 100 kb and 200 kb, and less than 5\% are larger than 200 kb (Fig. \ref{fig:pleiotropy_region_distribution}a).

% TODO continue here
The most pleiotropic region is 5:131,912,097-132,802,472 in cytoband 5q31.1 with a high concentration of cytokines genes (Table \ref{tab:pleiotropic_regions}).
%
The largest region is 11:13,260,511-17,396,930 in cytoband 11p15.2 with a length of 4,136,419 bp and variants involved in cardiovascular and hypertension traits (Supplementary table 5).
%
The second largest region is 2:187,235,912-191,066,738 in cytoband 2q32.2 with a length of 3,830,826 bp and variants involved in cardiovascular, hypertension and autoimmune traits  (Supplementary table 5).
